This lab on Ridge Regression and the Lasso in R comes from p. 251-255 of "Introduction to Statistical Learning with Applications in R" by Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani.

# 6.6: Ridge Regression and the Lasso

```{r}
library(ISLR)
library(glmnet)
```

We will use the ${\tt glmnet}$ package in order to perform ridge regression and
the lasso. The main function in this package is ${\tt glmnet()}$, which can be used
to fit ridge regression models, lasso models, and more. This function has
slightly different syntax from other model-fitting functions that we have
encountered thus far in this book. In particular, we must pass in an $x$
matrix as well as a $y$ vector, and we do not use the ${\tt y \sim x}$ syntax.

Before proceeding, let's first ensure that the missing values have
been removed from the data, as described in the previous lab.

```{r}
Hitters=na.omit(Hitters)
```

We will now perform ridge regression and the lasso in order to predict ${\tt Salary}$ on
the ${\tt Hitters}$ data. Let's set up our data:

```{r}
x=model.matrix(Salary~.,Hitters)[,-1]
y=Hitters$Salary
```

The ${\tt model.matrix()}$ function is particularly useful for creating $x$; not only
does it produce a matrix corresponding to the 19 predictors but it also
automatically transforms any qualitative variables into dummy variables.
The latter property is important because ${\tt glmnet()}$ can only take numerical,
quantitative inputs.

# 6.6.1 Ridge Regression
The ${\tt glmnet()}$ function has an alpha argument that determines what type
of model is fit. If ${\tt alpha=0}$ then a ridge regression model is fit, and if ${\tt alpha=1}$
then a lasso model is fit. We first fit a ridge regression model:

```{r}
grid=10^seq(10,-2,length=100)
ridge.mod=glmnet(x,y,alpha=0,lambda=grid)
```

By default the ${\tt glmnet()}$ function performs ridge regression for an automatically
selected range of $\lambda$ values. However, here we have chosen to implement
the function over a grid of values ranging from $\lambda = 10^10$ to $\lambda = 10^{-2}$, essentially
covering the full range of scenarios from the null model containing
only the intercept, to the least squares fit. As we will see, we can also compute
model fits for a particular value of $\lambda$ that is not one of the original
grid values. Note that by default, the ${\tt glmnet()}$ function standardizes the
variables so that they are on the same scale. To turn off this default setting,
use the argument ${\tt standardize=FALSE}$.

Associated with each value of $\lambda$ is a vector of ridge regression coefficients,
stored in a matrix that can be accessed by ${\tt coef()}$. In this case, it is a $20 \times 100$
matrix, with 20 rows (one for each predictor, plus an intercept) and 100
columns (one for each value of $\lambda$).

```{r}
dim(coef(ridge.mod))
```

We expect the coefficient estimates to be much smaller, in terms of $l_2$ norm,
when a large value of $\lambda$ is used, as compared to when a small value of $\lambda$ is
used. These are the coefficients when $\lambda = 11498$, along with their $l_2$ norm:

```{r}
ridge.mod$lambda[50] #Display 50th lambda value
coef(ridge.mod)[,50] # Display coefficients associated with 50th lambda value
sqrt(sum(coef(ridge.mod)[-1,50]^2)) # Calculate l2 norm
```

In contrast, here are the coefficients when $\lambda = 705$, along with their $l_2$
norm. Note the much larger $l_2$ norm of the coefficients associated with this
smaller value of $\lambda$.

```{r}
ridge.mod$lambda[60] #Display 60th lambda value
coef(ridge.mod)[,60] # Display coefficients associated with 60th lambda value
sqrt(sum(coef(ridge.mod)[-1,60]^2)) # Calculate l2 norm
```

We can use the ${\tt predict()}$ function for a number of purposes. For instance,
we can obtain the ridge regression coefficients for a new value of $\lambda$, say 50:

```{r}
predict(ridge.mod,s=50,type="coefficients")[1:20,]
```

We now split the samples into a training set and a test set in order
to estimate the test error of ridge regression and the lasso. There are two
common ways to randomly split a data set:
 - The first is to produce a random vector of ${\tt TRUE, FALSE}$ elements and select the observations corresponding to ${\tt TRUE}$ for the training data. 
 - The second is to randomly choose a subset of
numbers between $1$ and $n$; these can then be used as the indices for the
training observations. 
The two approaches work equally well. We used the
former method in the previous lab. Here we demonstrate the latter approach.
We first set a random seed so that the results obtained will be reproducible.

```{r}
set.seed(1)
train=sample(1:nrow(x), nrow(x)/2)
test=(-train)
y.test=y[test]
```

Next we fit a ridge regression model on the training set, and evaluate
its MSE on the test set, using $\lambda = 4$. Note the use of the ${\tt predict()}$
function again: this time we get predictions for a test set, by replacing
${\tt type="coefficients"}$ with the ${\tt newx}$ argument.

```{r}
ridge.mod=glmnet(x[train,],y[train],alpha=0,lambda=grid, thresh=1e-12)
ridge.pred=predict(ridge.mod,s=4,newx=x[test,])
mean((ridge.pred-y.test)^2)
```

The test MSE is 101037. Note that if we had instead simply fit a model
with just an intercept, we would have predicted each test observation using
the mean of the training observations. In that case, we could compute the
test set MSE like this:

```{r}
mean((mean(y[train])-y.test)^2)
```

We could also get the same result by fitting a ridge regression model with
a very large value of $\lambda$. Note that ${\tt 1e10}$ means $10^{10}$.

```{r}
ridge.pred=predict(ridge.mod,s=1e10,newx=x[test,])
mean((ridge.pred-y.test)^2)
```

So fitting a ridge regression model with $\lambda = 4$ leads to a much lower test
MSE than fitting a model with just an intercept. We now check whether
there is any benefit to performing ridge regression with $\lambda = 4$ instead of
just performing least squares regression. Recall that least squares is simply
ridge regression with $\lambda = 0$.

\* Note: In order for ${\tt glmnet()}$ to yield the **exact** least squares coefficients when $\lambda = 0$,
we use the argument ${\tt exact=T}$ when calling the ${\tt predict()}$ function. Otherwise, the
${\tt predict()}$ function will interpolate over the grid of $\lambda$ values used in fitting the
${\tt glmnet()}$ model, yielding approximate results. Even when we use ${\tt exact=T}$, there remains
a slight discrepancy in the third decimal place between the output of ${\tt glmnet()}$ when
$\lambda = 0$ and the output of ${\tt lm()}$; this is due to numerical approximation on the part of
${\tt glmnet()}$.

```{r}
ridge.pred=predict(ridge.mod,s=0,newx=x[test,],exact=T)
mean((ridge.pred-y.test)^2)
lm(y~x, subset=train)
predict(ridge.mod,s=0,exact=T,type="coefficients")[1:20,]
```

It looks like we are indeed improving over regular least-squares! Side note: in general, if we want to fit a (unpenalized) least squares model, then
we should use the ${\tt lm()}$ function, since that function provides more useful
outputs, such as standard errors and $p$-values for the coefficients.

Instead of arbitrarily choosing $\lambda = 4$, it would be better to
use cross-validation to choose the tuning parameter $\lambda$. We can do this using
the built-in cross-validation function, ${\tt cv.glmnet()}$. By default, the function
performs 10-fold cross-validation, though this can be changed using the
argument ${\tt folds}$. Note that we set a random seed first so our results will be
reproducible, since the choice of the cross-validation folds is random.

```{r}
set.seed(1)
cv.out=cv.glmnet(x[train,],y[train],alpha=0)          # Fit ridge regression model on training data
plot(cv.out)                                          # Draw plot of training MSE as a function of lambda
bestlam=cv.out$lambda.min                             # Select lamda that minimizes training MSE
```

Therefore, we see that the value of $\lambda$ that results in the smallest cross-validation
error is 212. What is the test MSE associated with this value of
$\lambda$?

```{r}
ridge.pred=predict(ridge.mod,s=bestlam,newx=x[test,]) # Use best lambda to predict test data
mean((ridge.pred-y.test)^2)                           # Calculate test MSE
```

This represents a further improvement over the test MSE that we got using
$\lambda = 4$. Finally, we refit our ridge regression model on the full data set,
using the value of $\lambda$ chosen by cross-validation, and examine the coefficient
estimates.

```{r}
out=glmnet(x,y,alpha=0)                           # Fit ridge regression model on full dataset
predict(out,type="coefficients",s=bestlam)[1:20,] # Display coefficients using lambda chosen by CV
```

As expected, none of the coefficients are exactly zero - ridge regression does not
perform variable selection!

# 6.6.2 The Lasso
We saw that ridge regression with a wise choice of $\lambda$ can outperform least
squares as well as the null model on the Hitters data set. We now ask
whether the lasso can yield either a more accurate or a more interpretable
model than ridge regression. In order to fit a lasso model, we once again
use the ${\tt glmnet()}$ function; however, this time we use the argument ${\tt alpha=1}$.
Other than that change, we proceed just as we did in fitting a ridge model:

```{r}
lasso.mod=glmnet(x[train,],y[train],alpha=1,lambda=grid) # Fit lasso model on training data
plot(lasso.mod)                                          # Draw plot of coefficients
```

Notice that in the coefficient plot that depending on the choice of tuning
parameter, some of the coefficients are exactly equal to zero. We now
perform cross-validation and compute the associated test error:

```{r}
set.seed(1)
cv.out=cv.glmnet(x[train,],y[train],alpha=1)          # Fit lasso model on training data
plot(cv.out)                                          # Draw plot of training MSE as a function of lambda
bestlam=cv.out$lambda.min                             # Select lamda that minimizes training MSE
lasso.pred=predict(lasso.mod,s=bestlam,newx=x[test,]) # Use best lambda to predict test data
mean((lasso.pred-y.test)^2)                           # Calculate test MSE
```

This is substantially lower than the test set MSE of the null model and of
least squares, and very similar to the test MSE of ridge regression with $\lambda$
chosen by cross-validation.

However, the lasso has a substantial advantage over ridge regression in
that the resulting coefficient estimates are sparse. Here we see that 12 of
the 19 coefficient estimates are exactly zero:

```{r}
out=glmnet(x,y,alpha=1,lambda=grid)                          # Fit lasso model on full dataset
lasso.coef=predict(out,type="coefficients",s=bestlam)[1:20,] # Display coefficients using lambda chosen by CV
lasso.coef
```

Selecting only the predictors with non-zero coefficients, we see that the lasso model with $\lambda$
chosen by cross-validation contains only seven variables:

```{r}
lasso.coef[lasso.coef!=0]                                    # Display only non-zero coefficients
```

To get credit for this lab, post your responses to the following questions:
 - How do ridge regression and the lasso improve on simple least squares?
 - In what cases would you expect ridge regression outperform the lasso, and vice versa?
 - What was the most confusing part of today's class?
 
to Piazza: https://piazza.com/class/igwiv4w3ctb6rg?cid=38