This lab on Model Validation using Validation and Cross-Validation in R comes from p. 248-251 of "Introduction to Statistical Learning with Applications in R" by Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani.

```{r}
library(ISLR)
library(leaps)
```

# Model selection using the Validation Set Approach

In Lab 8, we saw that it is possible to choose among a set of models of different
sizes using $C_p$, BIC, and adjusted $R^2$. We will now consider how to do this
using the validation set and cross-validation approaches.

As in Lab 8, we'll be working with the ${\tt Hitters}$ dataset from ${\tt ISLR}$. Since we're trying to predict ${\tt Salary}$ and we know from last time that some are missing, let's first drop all the rows with missing values:

```{r}
Hitters=na.omit(Hitters)
```

In order for these approaches to yield accurate estimates of the test
error, we must use *only the training observations* to perform all aspects of
model-fitting — including variable selection. Therefore, the determination of
which model of a given size is best must be made using *only the training
observations*. This point is subtle but important. If the full data set is used
to perform the best subset selection step, the validation set errors and
cross-validation errors that we obtain will not be accurate estimates of the
test error.

In order to use the validation set approach, we begin by splitting the
observations into a training set and a test set. We do this by creating
a random vector, train, of elements equal to TRUE if the corresponding
observation is in the training set, and FALSE otherwise. The vector test has
a TRUE if the observation is in the test set, and a FALSE otherwise. Note the
! in the command to create test causes TRUEs to be switched to FALSEs and
vice versa. We also set a random seed so that the user will obtain the same
training set/test set split.

```{r}
set.seed(1)
train=sample(c(TRUE,FALSE), nrow(Hitters),rep=TRUE)
test=(!train)
```

Now, we apply ${\tt regsubsets()}$ to the training set in order to perform best
subset selection\*.

(\* Note: If you're trying to complete this lab on a machine that can't handle **best subset**, try forward or backward selection instead by adding the ${\tt method = "forward"}$ or ${\tt method = "backward"}$ parameter to your call to ${\tt regsubsets()}$. You'll get slightly different values, but the concepts are the same.)

```{r}
regfit.best.train=regsubsets(Salary~.,data=Hitters [train ,], nvmax = 19)
```

Notice that we subset the ${\tt Hitters}$ data frame directly in the call in order
to access only the training subset of the data, using the expression
${\tt Hitters[train,]}$. We now compute the validation set error for the best
model of each model size. We first make a model matrix from the test
data.

```{r}
test.mat=model.matrix (Salary~.,data=Hitters [test ,])
```

The ${\tt model.matrix()}$ function is used in many regression packages for building an $X$ matrix from data. Now we run a loop, and for each size $i$, we
extract the coefficients from ${\tt regfit.best}$ for the best model of that size,
multiply them into the appropriate columns of the test model matrix to
form the predictions, and compute the test MSE.

```{r}
val.errors=rep(NA,19)

# Iterates over each size i
for(i in 1:19){
    
    # Extract the vector of predictors in the best fit model on i predictors
    coefi=coef(regfit.best.train,id=i)
    
    # Make predictions using matrix multiplication of the test matirx and the coefficients vector
    pred=test.mat[,names(coefi)]%*%coefi
    
    # Calculate the MSE
    val.errors[i]=mean((Hitters$Salary[test]-pred)^2)
}
```

Now let's plot the errors, and find the model that minimizes it:

```{r}
# Find the model with the smallest error
min = which.min(val.errors)

# Plot the errors for each model size
plot(val.errors,type='b')
points(min,val.errors[min][1], col ="red",cex =2, pch =20)
```

Viola! We find that the best model (according to the validation set approach) is the one that contains 10 predictors.

This was a little tedious, partly because there is no ${\tt predict()}$ method
for ${\tt regsubsets()}$. Since we will be using this function again, we can capture
our steps above and write our own ${\tt predict()}$ method:

```{r}
predict.regsubsets=function(object,newdata,id,...){
      form=as.formula(object$call[[2]]) # Extract the formula used when we called regsubsets()
      mat=model.matrix(form,newdata)    # Build the model matrix
      coefi=coef(object,id=id)          # Extract the coefficiants of the ith model
      xvars=names(coefi)                # Pull out the names of the predictors used in the ith model
      mat[,xvars]%*%coefi               # Make predictions using matrix multiplication
}
```

This function pretty much mimics what we did above. The one tricky
part is how we extracted the formula used in the call to ${\tt regsubsets()}$, but you don't need to worry too much about the mechanisc of this right now. We'll use this function to make our lives a little easier when we do cross-validation.

Now that we know what we're looking for, let's perform best subset selection on the full dataset (up to 10 predictors) and select the best 10-predictor model. It is important that we make use of the *full
data set* in order to obtain more accurate coefficient estimates. Note that
we perform best subset selection on the full data set and select the best 10-predictor
model, rather than simply using the predictors that we obtained
from the training set, because the best 10-predictor model on the full data
set may differ from the corresponding model on the training set.

```{r}
regfit.best=regsubsets(Salary~.,data=Hitters ,nvmax=10)
```

In fact, we see that the best ten-variable model on the full data set has a
**different set of predictors** than the best ten-variable model on the training
set:

```{r}
coef(regfit.best,10)
coef(regfit.best.train,10)
```

# Model selection using Cross-Validation

Now let's try to choose among the models of different sizes using cross-validation.
This approach is somewhat involved, as we must perform best
subset selection\* within each of the $k$ training sets. Despite this, we see that
with its clever subsetting syntax, ${\tt R}$ makes this job quite easy. First, we
create a vector that assigns each observation to one of $k = 10$ folds, and
we create a matrix in which we will store the results:

\* or forward selection / backward selection

```{r}
k=10        # number of folds
set.seed(1) # set the random seed so we all get the same results

# Assign each observation to a single fold
folds=sample(1:k,nrow(Hitters ), replace = TRUE)

# Create a matrix to store the results of our upcoming calculations
cv.errors=matrix(NA,k,19, dimnames=list(NULL, paste(1:19)))
```

Now let's write a for loop that performs cross-validation. In the $j^{th}$ fold, the
elements of folds that equal $j$ are in the test set, and the remainder are in
the training set. We make our predictions for each model size (using our
new $predict()$ method), compute the test errors on the appropriate subset,
and store them in the appropriate slot in the matrix ${\tt cv.errors}$.

```{r}
# Outer loop iterates over all folds
for(j in 1:k){
    
    # The perform best subset selection on the full dataset, minus the jth fold
    best.fit=regsubsets(Salary~.,data=Hitters[folds!=j,],nvmax=19)
    
    # Inner loop iterates over each size i
    for(i in 1:19){
        
        # Predict the values of the current fold from the "best subset" model on i predictors
        pred=predict(best.fit,Hitters[folds==j,],id=i)
        
        # Calculate the MSE, store it in the matrix we created above
        cv.errors[j,i]=mean((Hitters$Salary[folds==j]-pred)^2)
    }
}
```

This has filled up the ${\tt cv.errors}$ matrix such that the $(i,j)^{th}$ element corresponds
to the test MSE for the $i^{th}$ cross-validation fold for the best $j$-variable
model.  We can then use the ${\tt apply()}$ function to take the ${\tt mean}$ over the columns of this
matrix. This will give us a vector for which the $j^{th}$ element is the cross-validation
error for the $j$-variable model.

```{r}
# Take the mean of over all folds for each model size
mean.cv.errors = apply(cv.errors, 2, mean)

# Find the model size with the smallest cross-validation error
min = which.min(mean.cv.errors)

# Plot the cross-validation error for each model size, highlight the min
plot(mean.cv.errors,type='b')
points(min,mean.cv.errors[min][1], col ="red",cex =2, pch =20)
```

We see that cross-validation selects an 11-predictor model. Now let's use
best subset selection on the full data set in order to obtain the 11-predictor
model.

```{r}
reg.best=regsubsets(Salary~.,data=Hitters , nvmax =19)
coef(reg.best,11)
```

For comparison, let's also take a look at the statistics from last lab:

```{r}
par(mfrow=c(2,2))

reg.summary = summary(reg.best)

# Plot RSS
plot(reg.summary$rss,xlab="Number of Variables",ylab="RSS", type="l")

# Plot Adjusted R^2, highlight max value
plot(reg.summary$adjr2,xlab="Number of Variables", ylab=" Adjusted RSq",type="l")
max = which.max(reg.summary$adjr2)
points(max,reg.summary$adjr2[max], col ="red",cex =2, pch =20)

# Plot Cp, highlight min value
plot(reg.summary$cp,xlab="Number of Variables",ylab="Cp", type="l")
min = which.min(reg.summary$cp)
points(min,reg.summary$cp[min],col="red",cex=2,pch=20)

# Plot BIC, highlight min value
plot(reg.summary$bic,xlab="Number of Variables",ylab="BIC", type="l")
min = which.min(reg.summary$bic)
points(min,reg.summary$bic[min],col="red",cex=2,pch=20)
```

Notice how some of the indicators agree with the cross-validated model, and others are very different?

# Your turn!

Now it's time to test out these approaches (best / forward / backward selection) and evaluation methods (adjusted training error, validation set, cross validation) on other datasets. You may want to work with a team on this portion of the lab.

You may use any of the datasets included in the ${\tt ISLR}$ package, or choose one from the UCI machine learning repository (http://archive.ics.uci.edu/ml/datasets.html). Download a dataset, and try to determine the optimal set of parameters to use to model it!

```{r}
# Your code here
```

To get credit for this lab, please post your answers to the following questions:
 - What dataset did you choose?
 - Which selection techniques did you try?
 - Which evaluation techniques did you try?
 - What did you determine was the best set of parameters to model this data?
 - How well did this model perform?
 
to Piazza: https://piazza.com/class/igwiv4w3ctb6rg?cid=35

```{r}
```